Designing Types for R, Empirically

### 2020-05-16 paper/main.tex
stuff

@@

\paragraph{Structs}

In R, lists can have a names attribute, allowing lists elements to be
accessed by name using the \code{$} operator.

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
  agricolae::cv.model <- function(x) {
    suma2 <- sum(x$residual^2)
    gl <- x$df.residual
    promedio <- mean(x$fitted.values)
    return(sqrt(suma2/gl)*100/promedio)
  }
  data(sweetpotato)
  model<-aov(yield~virus, data=sweetpotato)
  cv.model(model)
\end{lstlisting}

In the code snippet, the function \code{cv.model} takes an argument
\code{x}, which we observed to always be a {\it linear model} (either ANOVA
with \CLASS{aov, lm} or a simple linear regression model with \CLASS{lm})
which is part of R's built-in statistics functionality.  Internally, linear
models are represented as lists with named elements, and one may access
these named elements using the \code{$} syntax, as seen on lines 2-4 in the
snippet.  To describe these kinds of values, we intended to include a struct
type in our type system, allowing users to specify name, type pairs as part
of list type definitions, though we found this to be problematic.

The inclusion of structs complicated our analysis by polluting the data,
illustrated on line 8 with the call \code{data(sweetpotato)}.  It is quite
common for R example code to use some built-in data sets, such as
\code{sweetpotato}, to show off some package functionality with ``real
data''.  When \typetracer is configured to detect structs we saw a marked
increase in misrepresentative types, as the types of functions which use
these data sets (such as \code{aov} in the code example) became polluted
with the names from the example sets.  We implemented a unification strategy
for struct types, removing them if they co-occurred with list types, and
turning them into lists if two co-occurring struct types shared no common
names, but even so we found ourselves with function signatures that were
full of noise introduced by the struct names.

We found that struct-like data was best described with classes, as
illustrated in the code snippet, and many structs actually had a class.
Even though \code{x} is used like a struct, it in fact has a linear model
class, which we believe is a more informative type.

@@

\paragraph{Objects}

R has multiple, disparate object systems, with the language providing direct
support for S3, S4, and R5.  Thus far, we've mainly discussed S3 and its
dispatch mechanism, and indeed our class types capture the S3
classes of values, but not user-defined polymorphism.  We will not discuss R5 as it is still under development.

S4, like S3, uses the class attribute to store types, we type them with a \CLASS{...} type.  
We stuck to simple class-based types for S4 since
doing anything more would be immensely complicated, and outside of the scope
of this work.  For instance, the mechanics of S4 dispatch are more complex
than for S3 (S4 dispatches on the class of {\it all} arguments), and users
can define their own class hierarchies that we would need to incorporate in
our type analysis and contract checking frameworks.  Further,we found
  limited use of S4 during our analysis. Coming up with a type system that
accounts for all of these factors and consolidates multiple
object-orientation frameworks in a single language design is an interesting
problem in and of itself, and we aim to extend our type annotation framework
to account for this in future work.  For now, simple class types for S3 and S4 suffice.

To illustrate informative class types, consider the following code snippet.
\code{distr} is a package which implements statistical distributions in an object-oriented way, and the \code{Binom} function acts as a constructor for \CLASS{Binom} values.
Another example is from the \code{Epi} package, which provides functionality for demographic and epidemiological analysis in the Lexis diagram.
The function we highlighted, \code{timeBand}, extracts time band data from the passed Lexis object.
We see from the type that the first argument is the Lexis object, and that either a \VEC\dbl or a \CLASS{factor} is returned.
Often, class members are implemented as attributes on R values, and it's likely that this is the case for Lexis objects as evidenced by the programmer accessing the "breaks" attribute in the function code.

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
#' @type <size: dbl, prob: dbl> => class<Binom>
distr::Binom <- function(size = 1,prob = 0.5) new("Binom", size = size, prob = prob)

#' @type <lex: class<`Lexis`, `data.frame`>, time.scale: chr, type: chr> => 
#'           class<`factor`> | dbl[];
Epi::timeBand <- function(lex, time.scale, type="integer") {
  time.scale <- check.time.scale(lex, time.scale)[1]
  breaks <- attr(lex, "breaks")[[time.scale]]
  ...
\end{lstlisting}

>>>>

\paragraph{Nullables}
The number of argument which may be \c{NULL} is 3,165 (or 3.5\%). This is a
relatively small number of occurrences, but it is worth expressing the
potential for the presence of \c{NULL} as these would likely inhibit
optimizations.

\paragraph{Structs}
While experimenting with various design, we consider adding a struct type to
capture lists with named elements that can be accessed with the \code{$}
operators. We ended discarding those types as they grew large and were often
only representative of the example data being manipulated.  Consider
function \code{cv.model}, its argument \code{x} is observed to be of
\CLASS{\texttt{aov}, \texttt{lm}} or \CLASS{\texttt{lm}}).  Internally,
linear models are represented as lists with named elements.  The polution is
illustrated by the lines after the funtion definition.  They load an example
data set and test the function \code{cv.model}.  The
\code{data(sweetpotato)} expression load a sample data set. The fields of
\code{sweetpotato} will be recorded when \code{cv.model} is called.

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
  cv.model <- function(x) {
    suma2 <- sum(x$residual^2)
    gl <- x$df.residual
    promedio <- mean(x$fitted.values)
    return(sqrt(suma2/gl)*100/promedio)
  }

  data(sweetpotato)
  model<-aov(yield~virus, data=sweetpotato)
  cv.model(model)
\end{lstlisting}

\paragraph{Objects}
While we record classes, our analysis does not deal with method dispatch.  R
has multiple, disparate, object systems, called S3, S4, and R5.  The
\code{class} attribute is used by these systems to dispatch methods. S3 does
single dispatch, S4 does multiple dispatch and R5 supports impertative
objects.  The mechanics of S4 dispatch are more complex than for S3, and
users can define their own class hierarchies that we would need to
incorporate in our type analysis and contract checking frameworks.  We found
limited use of S4 during our analysis. Coming up with a type system that
accounts for all of these factors and consolidates multiple
object-orientation frameworks in a single language design is an interesting
problem in and of itself, one we leave for future work.

@@

@@
